### 2020-02-04 manuscript/manuscript.rmd
word smithing

---
bibliography: manuscript.bib
csl: nature.csl
output:
  html_document: default
  word_document: default
  pdf_document: default
  github_document: default
---

@@

## Abstract
Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, a considerable debate still remains regarding the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult to resolve nodes in animal tree of life. Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses with these data in a consistent manner to characterize the differences between studies. We find that gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More importantly, we find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model and inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig. 1) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animal origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It has become clear that this is a very difficult phylogenetic challenge, and as the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary Results. Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyses; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models).

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT). If there are no modifiers, then it is implied that site homogeneous models are used. Moreover, the Gamma-rate heterogeinity is used in almost every analysis.

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document: default
  word_document: default
  html_document: default
  github_document: default
---

@@

## Abstract
Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool phylobayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig. 1) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animal origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). It has become clear that this is a very difficult phylogenetic challenge. As the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all phylogenomic analyses published befor 2018 that tested Ctenophora-sister and Porifera-sister, and for which we could obtain matrix partition data. We reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. This provides an integrative overview of the challenges posed by this question and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary Results. Briefly, the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices used in the studies under consideration here include Poisson (all rates equal; also known as F81), WAG and LG (fixed rate matrices derived from large databases of protein sequences), and GTR (rates inferred from the data at hand). The rate of excahnge can be further scaled, for example with a Gamma parameter that accommodate rate variation across sites. The approach to amino acid equilibrium frequencies includes using the frequencies observed in the data, estimating a single set of frequencies that are optimized over all sites for all taxa, or using heterogeneous models where frequencies are allowed to vary. Heterogeneous equilibrium approaches include CAT  [@Lartillot:2004dq], which is implemented in the software tool CAT and has been widely applied to questions of deep animal phylogeny, and the C models of iqtree that have until now not been applied to the problem.

Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. IN this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT). If there are no modifiers, then it is implied that site homogeneous models are used. Moreover, the Gamma-rate heterogeinity is used in almost every analysis.
